\Lecture{Jayalal Sarma}{October 14 2015}{40}{Proof of $R$ is a UFD implies
$R[x]$ is a UFD}{K Dinesh}{$\beta$}{K Dinesh}
Reference : \href{http://math.harvard.edu/~waffle/ufds2.pdf}{Notes by Waffle}

\Lecture{Jayalal Sarma}{October 16 2015}{41}{Characterization of irreducible
polynomials}{K Dinesh}{$\gamma$}{K Dinesh}

We saw in last two lectures that the following problems are well defined.
\begin{problem}
	Given a polynomial $f \in \polyring$, 
	\begin{itemize}
		\item test if it is irreducible or not,
		\item if not, factorize $f$ into irreducible factors.
	\end{itemize}
\end{problem}

In this lecture, we give a characterization of irreducible polynomials over
$\F[x]$. Recall the following definitions.

\begin{definition}[Quotient Ring]
	For an ideal $I$ in a ring $R(+,\cdot)$, the quotient ring $(R/I, +
	,\cdot)$ is a ring defined as the collection of $\{a+I ~|~ a \in R\}$
	with operations defined as : for all $a+I, b+I \in R \ I$,
	$(a+I)+(b+I) = (a+b+ I)$ and $(a+I) \cdot (b+I) = a\cdot b+I$.
\end{definition}
The element of $R/I$ are the cosets of $I$ in $R$ and this forms a ring for
the operations defined.  Following lemma gives as a characterization of when
do two elements in $R$ fall in the same coset.
\begin{lemma} \label{lem:mem-ideal}
	For an ideal $I$ of a ring $R$, for any $a,b \in R$ 
	\[ a+I = b + I \iff a-b \in I \]
\end{lemma}
\begin{proof}
	($\Longrightarrow$) Let $a+I = b+I$. In particular $a \in a+I$. Hence
	$a \in b+I$ and $a-b \in I$.

	($\Longleftarrow$) We show that if $a-b \in I$, then $a+I \subseteq
	b+I$. Let $a-b = \alpha$ for some $\alpha \in I$. Hence $a = b +
	\alpha$ which implies $a \in b+I$. Now for any $c \in a+I$, $c \in
	b+(a-b)+I$. Since $a-b = \alpha \in I$, $(a-b)+I = I$ giving $c \in
	b+I$. Now interchange the role of $a$ and $b$ in the above proof to
	get $b+I \subseteq a+I$. This completes the proof.
\end{proof}

Let us consider an example of quotient rings. As usual, consider the ring $\Z$
and let $a\Z$ for $a \in \Z$ be an ideal. Recall that $\Z$ is a PID and hence
every ideal is singly generated. Consider the quotient ring $\Z/a\Z$. 
By earlier lemma (Lemma~\ref{lem:mem-ideal}), we get that 
\begin{observation} \label{obs:mem-z}
	For any $\alpha + I$, $\beta + I \in \Z/a\Z$, $\alpha$ and $\beta$
	are in the same cosets if and only if $\alpha - \beta \in a\Z$. 
\end{observation}
This is essentially same as that of saying $\alpha = \beta \mod a$. 
Note that there cannot be two representatives for any coset. If $\alpha,
\beta$ represent the same coset $a\Z$ with $\alpha \ne \beta$, then
$\alpha-\beta \ne 0 \mod a$ which violates the above observation
(Observation~\ref{obs:mem-z}).

Also coset representatives is of the coset are those that are less than
$a$ in value. If not, then we can always go modulo $a$ and get a
representative smaller than $a$. Hence the set quotient ring $\Z/a\Z$ can be
looked as $\Z_a$.
\begin{claim}
	$\Z/a\Z$ is isomorphic to $\Z_a$
\end{claim}
\begin{proof}
	For any $\alpha + a\Z \in \Z/a\Z$, define the map $\phi$ which takes
	$\alpha + a\Z$ to $\alpha \mod a$. This is indeed a bijection.  
	Also the operation is preserved because,
	\begin{align*}
		\phi(\alpha + a\Z + \beta +b\Z) = & \phi(\alpha + \beta +a\Z)
		= (\alpha + \beta) \mod a \\ 
		& = (\alpha \mod a + \beta \mod a ) \mod a \\
		& = \phi(\alpha + a\Z) + \phi(\beta +a\Z)
	\end{align*}
\end{proof}

\section{Quotient rings over $\F[x]$}
For a polynomial $p(x) \in \F[x]$ with $deg(p) = d$, consider the ideal $I =
\langle p(x) \rangle$ and the quotient ring $\F[x]/I$. The multiplicative 
identity is $1 + I$ and additive identity is $0 + I$.  Similar to the integer
rings, we make the following observations.

\begin{observation}
	No two polynomials with degree strictly smaller than $deg(p)$ can be
	in the same coset. 
\end{observation}
This follows from Lemma~\ref{lem:mem-ideal} and the argument is similar to the
integer case. Hence every coset has at most one polynomial of degree strictly
smaller in it. The next observation says that every coset must have an element
of degree strictly smaller than $deg(p)$.

\begin{observation}
	Every coset in $\F[x]/I$ must have a polynomial of degree strictly
	smaller than $deg(p)$.
\end{observation}
\begin{proof}
	Let $g + I$ be a coset in $\F[x]/I$. By division algorithm, $g(x)$ can
	be written as $q(x)p(x) + r(x)$ with $deg(r) < deg(p)$ and $r$ is
	unique. Hence $r(x) = g(x) - p(x)q(x) \in g +I$.
\end{proof}

This tells us that corresponding to every polynomial of degree strictly
smaller than $deg(p)$, there is exactly one coset. Hence for a finite field
$\F$ of size $q$, $|\F[x]/I| = q^d$.  We now give a characterization of
irreducible polynomials.  
\section{Characterization of Irreducibility}
The characterization essentially says that checking for irreducibility holds
down to checking if a ring is a field.
\begin{theorem}
	For an ideal $I$ in $\F[x]$ generated by $p(x) \in \F[x]$, 
	\[ \F[x]/I \text{ is a field} \iff p(x) \text{ is irreducible over }
	\F[x] \]
\end{theorem}
\begin{proof}
	($\Longleftarrow$) Suppose $p(x)$ is irreducible in $\F[x]$, we need
	to show that $\F[x]/I$ is a field. Since we already know that
	$\F[x]/I$ is a ring, with $1+p(x)$ as the multiplicative identity it
	suffices to show that for all elements $g + I \in \F[x]/I$ with $g
	\not \equiv 0$ and not a constant, $g$ has an inverse. 

	Since $g$ is non-constant and $p$ is irreducible $gcd(p,g) = 1$. Since
	$\langle p(x), g(x) \rangle = \langle gcd(p,g) \rangle$
	(Lemma~\ref{lem:gcd}), we get that $\langle p(x) ,g(x) \rangle =
	\langle 1 \rangle$. Hence there exists $\alpha(x), \beta(x) \in \F[x]$
	such that $\alpha(x)p(x) + \beta(x)g(x) = 1$. Hence $g(x)\beta(x) = 1
	- \alpha(x)p(x)$. Hence $g(x) \beta(x) = 1 + I$. Hence $\beta(x)$ is
	the inverse.

	($\Longleftarrow$) We show that if $p(x)$ is not irreducible in
	$\F[x]$, then we show that $\F[x]/I$ is not even an integral domain. 

	Suppose $p(x)$ is not irreducible and $p_1(x)$ and $p_2(x)$ are two
	non-constant factors such that $p(x) = p_1(x)p_2(x)$. Now
	consider the two elements $p_1(x) + I$ and $p_2 + I$ in $\F[x]/I$.
	Then $(p_1(x) + I) (p_2(x) + I) = p_1(x)p_2(x) + I = 0 + I$. Hence the
	ring is not an integral domain.
\end{proof}



